---
title: "Large-Sample Normal Approximation to the Posterior"
subtitle: "First Observaton: for large n, the prior doesn't matter"
output:
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{multicol}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(gridExtra)
```

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

# Poisson Model

 * Observe $X_1, \ldots, X_n$; $X_i$ is the number of seedlings in quadrat number $i$.
 * Data Model: $X_i | \Lambda = \lambda \simiid \text{Poisson}(\lambda)$
 * Suppose we use a Gamma prior for $\Lambda$
    * Example: $\Lambda \sim \text{Gamma}(1, 0.01)$ is fairly non-informative
 * Think of the posterior pdf in terms of contributions from the prior and the likelihood


\newpage

## Simulation: Suppose $\lambda = 10$

#### n = 1

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

plot_xmax <- 50

n <- 1
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 6}
prior_shape <- 1
prior_rate <- 1 # 0.01

calc_ind_lik <- function(lambda, x, log = TRUE) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  
  if(log) {
    return(result)
  } else {
    return(exp(result))
  }
}

calc_post_plus_lik <- function(lambda, x, log = TRUE) {
  temp <- dgamma(lambda, shape = prior_shape, rate = prior_rate, log = TRUE) +
    calc_ind_lik(lambda, x)
  
  if(log) {
    return(temp)
  } else {
    return(exp(temp))
  }
}


plot_df <- data.frame(
  lambda = seq(from = 0, to = plot_xmax, length = 1001)
) %>%
  mutate(
    Prior = dgamma(lambda, shape = prior_shape, rate = prior_rate, log = FALSE),
    Likelihood = calc_ind_lik(lambda, x = x_sim$x, log = FALSE),
    Posterior = dgamma(lambda, shape = prior_shape + sum(x_sim$x), rate = prior_rate + n, log = FALSE),
    log_Prior = dgamma(lambda, shape = prior_shape, rate = prior_rate, log = TRUE),
    log_Likelihood = calc_ind_lik(lambda, x = x_sim$x, log = TRUE),
    log_Posterior = dgamma(lambda, shape = prior_shape + sum(x_sim$x), rate = prior_rate + n, log = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = c("Prior", "Likelihood", "Posterior", "log_Prior", "log_Likelihood", "log_Posterior"),
    names_to = "Curve",
    values_to = "value"
  ) %>%
  mutate(
    Curve = factor(Curve, levels = c("Prior", "Likelihood", "Posterior", "log_Prior", "log_Likelihood", "log_Posterior"))
  )

p1 <- ggplot(data = plot_df %>% filter(Curve %in% c("Prior", "Likelihood", "Posterior"))) +
  geom_line(mapping = aes(x = lambda, y = value, color = Curve, linetype = Curve)) +
  scale_color_viridis_d("Curve", begin = 0.2, end = 0.7, option = "D") +
  scale_linetype_discrete("Curve") +
  theme_bw() +
  facet_wrap( ~ Curve, nrow = 1, scales = "free_y") +
  ggtitle("Prior, Likelihood, and Posterior - Different Vertical Scales")

p2 <- ggplot(data = plot_df %>% filter(Curve %in% c("log_Prior", "log_Likelihood", "log_Posterior"))) +
  geom_line(mapping = aes(x = lambda, y = value, color = Curve, linetype = Curve)) +
  scale_color_viridis_d("Curve", begin = 0.2, end = 0.7, option = "D") +
  scale_linetype_discrete("Curve") +
  theme_bw() +
  ggtitle("Log Prior, Log Likelihood, and Log Posterior - Same Vertical Scale")

grid.arrange(p1, p2, ncol = 1)
```

\newpage

#### n = 100

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

plot_xmax <- 50

n <- 100
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```


```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 6}
prior_shape <- 1
prior_rate <- 1 # 0.01

calc_ind_lik <- function(lambda, x, log = TRUE) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  
  if(log) {
    return(result)
  } else {
    return(exp(result))
  }
}

calc_post_plus_lik <- function(lambda, x, log = TRUE) {
  temp <- dgamma(lambda, shape = prior_shape, rate = prior_rate, log = TRUE) +
    calc_ind_lik(lambda, x)
  
  if(log) {
    return(temp)
  } else {
    return(exp(temp))
  }
}


plot_df <- data.frame(
  lambda = seq(from = 0, to = plot_xmax, length = 1001)
) %>%
  mutate(
    Prior = dgamma(lambda, shape = prior_shape, rate = prior_rate, log = FALSE),
    Likelihood = calc_ind_lik(lambda, x = x_sim$x, log = FALSE),
    Posterior = dgamma(lambda, shape = prior_shape + sum(x_sim$x), rate = prior_rate + n, log = FALSE),
    log_Prior = dgamma(lambda, shape = prior_shape, rate = prior_rate, log = TRUE),
    log_Likelihood = calc_ind_lik(lambda, x = x_sim$x, log = TRUE),
    log_Posterior = dgamma(lambda, shape = prior_shape + sum(x_sim$x), rate = prior_rate + n, log = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = c("Prior", "Likelihood", "Posterior", "log_Prior", "log_Likelihood", "log_Posterior"),
    names_to = "Curve",
    values_to = "value"
  ) %>%
  mutate(
    Curve = factor(Curve, levels = c("Prior", "Likelihood", "Posterior", "log_Prior", "log_Likelihood", "log_Posterior"))
  )

p1 <- ggplot(data = plot_df %>% filter(Curve %in% c("Prior", "Likelihood", "Posterior"))) +
  geom_line(mapping = aes(x = lambda, y = value, color = Curve, linetype = Curve)) +
  scale_color_viridis_d("Curve", begin = 0.2, end = 0.7, option = "D") +
  scale_linetype_discrete("Curve") +
  theme_bw() +
  facet_wrap( ~ Curve, nrow = 1, scales = "free_y") +
  ggtitle("Prior, Likelihood, and Posterior - Different Vertical Scales")

p2 <- ggplot(data = plot_df %>% filter(Curve %in% c("log_Prior", "log_Likelihood", "log_Posterior"))) +
  geom_line(mapping = aes(x = lambda, y = value, color = Curve, linetype = Curve)) +
  scale_color_viridis_d("Curve", begin = 0.2, end = 0.7, option = "D") +
  scale_linetype_discrete("Curve") +
  theme_bw() +
  ggtitle("Log Prior, Log Likelihood, and Log Posterior - Same Vertical Scale")

grid.arrange(p1, p2, ncol = 1)
```
