---
title: "Definitions and Example for Confidence Intervals"
output: pdf_document
header-includes:
   - \usepackage{afterpage}
---

\afterpage{\null\newpage\null\newpage}

\newpage

# Example: Body Temperatures

* It's generally believed that the average body temperature is 98.6 degrees Farenheit (37 degrees Celsius).
* Let's investigate with measurements of the temperatures of 130 adults.

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
require(ggplot2)
require(dplyr)
bodytemp = read.table('http://www.amstat.org/publications/jse/datasets/normtemp.dat.txt')
names(bodytemp) <- c('temp', 'sex', 'hr')
```

### Confidence Interval Calculation

Sample mean and standard deviation:

```{r}
x_bar <- mean(bodytemp$temp)
x_bar
s <- sd(bodytemp$temp)
s
```

Verifying our sample size is $n = 130$

```{r}
n <- nrow(bodytemp)
n
```

Finding appropriate quantiles for a 95% confidence interval:

```{r}
t_lower <- qt(0.025, df = n - 1)
t_lower
t_upper <- qt(0.975, df = n - 1)
t_upper
```

Calculation of the confidence interval:

```{r}
x_bar - t_upper * s / sqrt(n)
x_bar - t_lower * s / sqrt(n)
```





