---
title: "dplyr::mutate and dplyr::summarize"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{booktabs}
geometry: margin=1.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r, echo = FALSE}
iris_short <- iris[1:5, ]
```

### Running example

So that we have a running example to work with, here are the first 5 rows of the `iris` data set that comes with R:

```{r}
iris_short
```

### Logical Operations on Vectors

We can compare vectors to each other using:

 * `<` less than
 * `<=` less than or equal to
 * `==` equal to
 * `>=` greater than or equal to
 * `>` greater than
 * `!=` not equal to

```{r}
iris_short$Sepal.Length > iris_short$Sepal.Width
iris_short$Sepal.Length >= 5.0
iris_short$Sepal.Length != 5.0
```

We can combine multiple conditions using:

 * `&` is TRUE if both conditions are TRUE
 * `|` is TRUE if either (or both) condition is TRUE

```{r}
iris_short$Sepal.Length > 4.6 & iris_short$Sepal.Length < 5.0
iris_short$Sepal.Length <= 4.6 | iris_short$Sepal.Length >= 5.0
```

When you perform arithmetic operations on a logical vector, `TRUE` is converted to 1 and `FALSE` is converted to 0:

```{r}
sum(iris_short$Sepal.Length <= 4.6 | iris_short$Sepal.Length >= 5.0)
mean(iris_short$Sepal.Length <= 4.6 | iris_short$Sepal.Length >= 5.0)
```

#### Defining a data frame

The format of the `mutate` and `summarize` commands is less confusing if we have the way to create a data frame firmly in mind:

```{r}
example_df <- data.frame(
  a = c("v", "w", "x", "y", "z"),
  b = 10 * sqrt(1:5)
)

example_df
```

#### `dplyr` package

The `mutate` and `summarize` functions are provided by the `dplyr` package.

```{r}
library(dplyr)
```

#### `mutate`: add a new variable or modify an existing variable in a data frame

 * First argument: name of data frame to modify
 * Remaining arguments: variables and how to compute them.  Refer to variables in the data frame directly.

```{r}
mutate(iris_short,
  a = c("v", "w", "x", "y", "z"),
  b = 10 * sqrt(1:5),
  sepal_sum = Sepal.Length + Sepal.Width
)
```

 * The original data frame is not changed, a copy is made
 * The pipe `%>%` calls the function on the right using the expresson on the left as the first argument

```{r}
iris_updated <- iris_short %>% mutate(
  a = c("v", "w", "x", "y", "z"),
  b = 10 * sqrt(1:5),
  sepal_sum = Sepal.Length + Sepal.Width
)

iris_updated

iris_short
```

#### `summarize`: compute length-1 summaries of a data frame

```{r}
mean(iris_short$Sepal.Length)
sd(iris_short$Sepal.Length)

iris_short %>%
  summarize(
    mean_sepal_length = mean(Sepal.Length),
    sd_sepal_length = sd(Sepal.Length)
  )
```

 * Longer-length summaries result in errors:

```{r, error = TRUE}
quantile(iris_short$Sepal.Length)

iris_short %>%
  summarize(
    quantiles_sepal_length = quantile(Sepal.Length)
  )
```
