---
title: "Likelihood Ratios"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
geometry: margin=0.6in
---

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

 * Data Model: $X_1, \ldots, X_5 \simiid \text{Normal}(\theta, 5^2)$

 * Let's consider a test of the hypotheses $H_0: \theta = 25$ vs. $H_A: \theta = 20$

 * If $H_0$ is correct, then $\bar{X} \sim \text{Normal}(25, 5^2/5)$.  If $H_A$ is correct, then $\bar{X} \sim \text{Normal}(20, 5^2/5)$

 * Two ways to think of the specification of the rejection region for the likelihood ratio test:

#### 1. Reject $H_0$ if $\bar{x} \leq 21.322$

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 2, fig.width = 7, fig.align = 'center'}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5)), color = "cornflowerblue") +
  stat_function(fun = dnorm, args = list(mean = 20, sd = sqrt(5)), color = "orange") +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("\n density")
```

#### 2. Reject $H_0$ if $W \leq w^*$ where $W$ is the likelihood ratio statistic

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 2, fig.width = 7, fig.align = 'center'}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

lr <- function(x) {
  dnorm(x, mean = 25, sd = sqrt(5)) / dnorm(x, mean = 20, sd = sqrt(5))
}

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = lr) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("likelihood ratio, w")
```

The scale of the likelihood ratio is challenging; let's look at the log:

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 2, fig.width = 7, fig.align = 'center'}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

lr <- function(x) {
  dnorm(x, mean = 25, sd = sqrt(5), log = TRUE) - dnorm(x, mean = 20, sd = sqrt(5), log = TRUE)
}

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = lr) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("\n log(likelihood ratio)")
```


