---
title: "Warm Up: Power Functions for Hypothesis Tests"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
geometry: margin=0.6in
---

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

 * Data Model: $X_1, \ldots, X_5 \simiid \text{Normal}(\theta, 5^2)$.  We saw that the likelihood ratio test is *equivalent* to a test based on $\bar{x}$.

#### 1. Consider a test of the hypotheses $H_0: \theta = 25$ vs. $H_A: \theta = 20$.  Below is a picture showing the pdf $f_{\bar{X}|\theta}(\bar{x} | 20)$ of a $\text{Normal}(20, 5^2/5)$ distribution and the pdf $f_{\bar{X}|\theta}(\bar{x} | 25)$ of a $\text{Normal}(25, 5^2/5)$ distribution, along with a vertical line at $q_5^{null}$, the 5th percentile of the $\text{Normal}(25, 5^2/5)$ distribution.

 * **Shade in the area corresponding to $1 - \beta$, the power of the likelihood ratio test if $H_A$ is correct.**
 * **Show how you would calculate the power of the test as an integral of either $f_{\bar{X}|\theta}(\bar{x} | 20)$ or $f_{\bar{X}|\theta}(\bar{x} | 25)$**

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.25, fig.width = 7}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 20, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1,
    alpha = 0.3) +  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```

$1 - \beta = \int_{-\infty}^{q_5^{null}}f_{\bar{X}|\theta}(\bar{x} | 20) d \bar{x}$

#### 2. Suppose that instead we were testing the hypotheses $H_0: \theta = 25$ vs. $H_A: \theta = 22.5$. Below is a picture showing the pdf $f_{\bar{X}|\theta}(\bar{x} | 22.5)$ of a $\text{Normal}(22.5, 5^2/5)$ distribution and the pdf $f_{\bar{X}|\theta}(\bar{x} | 25)$ of a $\text{Normal}(25, 5^2/5)$ distribution, along with a vertical line at $q_5^{null}$, the 5th percentile of the $\text{Normal}(25, 5^2/5)$ distribution.

 * **Shade in the area corresponding to $1 - \beta$, the power of the likelihood ratio test if $H_A$ is correct.**
 * **Show how you would calculate the power of the test as an integral of either $f_{\bar{X}|\theta}(\bar{x} | 22.5)$ or $f_{\bar{X}|\theta}(\bar{x} | 25)$**

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.25, fig.width = 7}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 22.5, sd = sqrt(5), log = FALSE), 0)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 22.5, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1,
    alpha = 0.3) +  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```

$1 - \beta = \int_{-\infty}^{q_5^{null}}f_{\bar{X}|\theta}(\bar{x} | 22.5) d \bar{x}$

#### 3. Suppose that instead we were testing the hypotheses $H_0: \theta = 25$ vs. $H_A: \theta = 17.5$. Below is a picture showing the pdf $f_{\bar{X}|\theta}(\bar{x} | 17.5)$ of a $\text{Normal}(17.5, 5^2/5)$ distribution and the pdf $f_{\bar{X}|\theta}(\bar{x} | 25)$ of a $\text{Normal}(25, 5^2/5)$ distribution, along with a vertical line at $q_5^{null}$, the 5th percentile of the $\text{Normal}(25, 5^2/5)$ distribution.

 * **Shade in the area corresponding to $1 - \beta$, the power of the likelihood ratio test if $H_A$ is correct.**
 * **Show how you would calculate the power of the test as an integral of either $f_{\bar{X}|\theta}(\bar{x} | 17.5)$ or $f_{\bar{X}|\theta}(\bar{x} | 25)$**

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.25, fig.width = 7}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 17.5, sd = sqrt(5), log = FALSE), 0)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 17.5, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1,
    alpha = 0.3) +  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```

$1 - \beta = \int_{-\infty}^{q_5^{null}}f_{\bar{X}|\theta}(\bar{x} | 17.5) d \bar{x}$

#### 4. For which of the alternative hypotheses above ($\theta = 17.5$, $\theta = 20$, or $\theta = 22.5$) is the power of the test largest?  For which is the power smallest?

The power is largest for $\theta = 17.5$ and smallest for $\theta = 22.5$.

\newpage

#### Definition: power function

The *power function* $K(\theta)$ for a test is the power of the test at $\theta$:

$K(\theta) = \int_{R}f_{\mathbf{X}|\theta}(\mathbf{x} | \theta) d \mathbf{x}$

where $R$ denotes the rejection region of the test (i.e., $R$ is the set of $\mathbf{x}$ such that the p-value is less than $\alpha$)

#### Definition: Uniformly most powerful test

A test with power function $K(\theta)$ if for any other test with power function $K'(\theta)$, $K(\theta) \geq K'(\theta)$ for all $\theta \in \Omega_0^c$ (i.e., it has at least as large of power as any other test for every possible parameter consistent with the alternative hypothesis).

#### (Non-) Example

Here is a plot of the power functions of 3 different tests of

$H_0: \theta = 25$ vs. $H_A: \theta \neq 25$

for our running example with $X_1, \ldots, X_n \simiid \text{Normal}(\theta, 5^2)$.

 * Shown in black is the power function for a test that rejects $H_0$ if $\bar{x} \leq q_{0.025}^{null}$ or $q_{0.975}^{null} \leq \bar{x}$
 * Shown in blue is the power function for a test that rejects $H_0$ if $\bar{x} \leq q_{0.05}^{null}$
 * Shown in orange is the power function for a test that rejects $H_0$ if $q_{0.95}^{null} \leq \bar{x}$

```{r, fig.height = 2, fig.width = 7, echo = TRUE}
calc_power_two_sided <- function(theta) {
  lower_cutoff <- qnorm(0.025, mean = 25, sd = sqrt(5))
  upper_cutoff <- qnorm(0.975, mean = 25, sd = sqrt(5))
  
  power <- rep(NA, length(theta))
  for(i in seq_along(theta)) {
    power[i] <- pnorm(lower_cutoff, mean = theta[i], sd = sqrt(5)) +
      pnorm(upper_cutoff, mean = theta[i], sd = sqrt(5), lower.tail = FALSE)
  }
  return(power)
}

calc_power_left_side <- function(theta) {
  lower_cutoff <- qnorm(0.05, mean = 25, sd = sqrt(5))

  power <- rep(NA, length(theta))
  for(i in seq_along(theta)) {
    power[i] <- pnorm(lower_cutoff, mean = theta[i], sd = sqrt(5))
  }
  return(power)
}

calc_power_right_side <- function(theta) {
  upper_cutoff <- qnorm(0.95, mean = 25, sd = sqrt(5))

  power <- rep(NA, length(theta))
  for(i in seq_along(theta)) {
    power[i] <- pnorm(upper_cutoff, mean = theta[i], sd = sqrt(5), lower.tail = FALSE)
  }
  
  return(power)
}

ggplot(data = data.frame(theta = c(0, 50)), mapping = aes(x = theta)) +
  stat_function(fun = calc_power_two_sided) +
  stat_function(fun = calc_power_left_side, color = "cornflowerblue") +
  stat_function(fun = calc_power_right_side, color = "orange") +
  ylim(0, 1) +
  geom_hline(yintercept = 0.05, linetype = 2) +
  theme_bw()
```

 * By the Neyman-Pearson lemma, the left-sided test (blue) is most powerful for any $\theta_1 < \theta_0$
 * By the Neyman-Pearson lemma, the right-sided test (orange) is most powerful for any $\theta_1 > \theta_0$
 * Therefore, there is no test that is uniformly most powerful for a two-sided test about the mean of a normal distribution.
