---
title: "Warm Up: Motivation for the Neyman-Pearson Lemma"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
geometry: margin=0.6in
---

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

 * Data Model: $X_1, \ldots, X_5 \simiid \text{Normal}(\theta, 5^2)$

 * Let's consider a test of the hypotheses $H_0: \theta = 25$ vs. $H_A: \theta = 20$

 * If $H_0$ is correct, then $\bar{X} \sim \text{Normal}(25, 5^2/5)$. If $H_A$ is correct, then $\bar{X} \sim \text{Normal}(20, 5^2/5)$

#### 1. The following pictures can be used to illustrate 3 different tests based on the sampling distribution of $\bar{X}$, all with P(Type I Error | $H_0$ true) = 0.05.  For each test,

 * **Shade in the area corresponding to the probability of a Type I Error** (blue)
 * **Shade in the area corresponding to the power of the test** (orange)

Test 1: Reject $H_0$ if $\bar{x} \leq 21.322$ (This is the likelihood ratio test: for values of $\bar{x} \leq 21.322$, the p-value is $\leq 0.05$.)

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.5, fig.width = 7}
x_grid_1 <- seq(from = 0, to = qnorm(0.05, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

region_to_shade1a <- data.frame(
  x = c(0, x_grid_1, qnorm(0.05, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 20, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1,
    alpha = 1) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1a,
    alpha = 0.3) +  geom_vline(xintercept = qnorm(0.05, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```

Test 2: Reject $H_0$ if $\bar{x} \leq 20.617$ or $\bar{x} \geq 29.383$

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.5, fig.width = 7}
x_grid_1 <- seq(from = 0, to = qnorm(0.025, mean = 25, sd = sqrt(5)), length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, qnorm(0.025, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

region_to_shade1a <- data.frame(
  x = c(0, x_grid_1, qnorm(0.025, mean = 25, sd = sqrt(5))),
  y = c(0, dnorm(x_grid_1, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

x_grid_2 <- seq(from = qnorm(0.975, mean = 25, sd = sqrt(5)), to = 35, length = 101)

region_to_shade2 <- data.frame(
  x = c(qnorm(0.975, mean = 25, sd = sqrt(5)), x_grid_2, 35),
  y = c(0, dnorm(x_grid_2, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

region_to_shade2a <- data.frame(
  x = c(qnorm(0.975, mean = 25, sd = sqrt(5)), x_grid_2, 35),
  y = c(0, dnorm(x_grid_2, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 20, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade2) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1a,
    alpha = 0.3) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade2a,
    alpha = 0.3) +
  geom_vline(xintercept = qnorm(0.025, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.975, mean = 25, sd = sqrt(5))) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```


Test 3: Reject $H_0$ if $20.617 \leq \bar{x} \leq 21.322$ or $28.678 \leq \bar{x} \leq 29.383$

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.5, fig.width = 7}
x_grid_1 <- seq(from = 20.617, to = 21.322, length = 101)

region_to_shade1 <- data.frame(
  x = c(x_grid_1[1], x_grid_1, tail(x_grid_1, 1)),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

region_to_shade1a <- data.frame(
  x = c(x_grid_1[1], x_grid_1, tail(x_grid_1, 1)),
  y = c(0, dnorm(x_grid_1, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

x_grid_2 <- seq(from = 28.678, to = 29.383, length = 101)

region_to_shade2 <- data.frame(
  x = c(x_grid_2[1], x_grid_2, tail(x_grid_2, 1)),
  y = c(0, dnorm(x_grid_2, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

region_to_shade2a <- data.frame(
  x = c(x_grid_2[1], x_grid_2, tail(x_grid_2, 1)),
  y = c(0, dnorm(x_grid_2, mean = 20, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

ggplot(data = data.frame(x = c(0, 35)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  stat_function(fun = dnorm, args = list(mean = 20, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.025, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.025 + 0.025, mean = 25, sd = sqrt(5))) +
#  geom_vline(xintercept = qnorm(0.5 - 0.0125, mean = 25, sd = sqrt(5))) +
#  geom_vline(xintercept = qnorm(0.5 + 0.0125, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.975 - 0.025, mean = 25, sd = sqrt(5))) +
  geom_vline(xintercept = qnorm(0.975, mean = 25, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade2) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade1a,
    alpha = 0.3) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "orange",
    data = region_to_shade2a,
    alpha = 0.3) +
  coord_cartesian(xlim = c(10, 35), expand = TRUE) +
  theme_bw() +
  xlab(expression(bar(x))) +
  ylab("density")
```

#### 2. Which of the tests above has the highest power?

The likelihood ratio test.

#### 3. For the likelihood ratio test, write down how you would calculate the probability of making a Type I Error and the power of the test as suitable integrals of either $f_{\bar{X}|\theta}(\bar{x} | 20)$ or $f_{\bar{X}|\theta}(\bar{x} | 25)$.  (You will have 1 integral for the probability of a Type I Error and a second for the power of the test.)

$P(\text{Type I Error} | H_0 \text{ correct}) = \int_{-\infty}^{21.322}f_{\bar{X}|\theta}(\bar{x} | 25) d \bar{x}$

$Power = 1 - \beta = \int_{-\infty}^{21.322}f_{\bar{X}|\theta}(\bar{x} | 20) d \bar{x}$
