---
title: "Hypothesis Testing - Finishing the Paint Drying Example"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
geometry: margin=0.6in
---

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

 * Data Model: $X_1, \ldots, X_5 \simiid \text{Normal}(\theta, 5^2)$

 * We know that $\theta$ is either 10 (quick dry) or 25 (regular).

 * We have a vague feeling that batch 1 is quick-dry; we will calculate a p-value to see how strong the evidence is that the batch we used is *not* 25 minutes.

$H_0: \theta = 25$ vs. $H_A: \theta = 10$

 * After collecting data, we observe a sample mean of $\bar{x} = 20$ minutes.

 * We saw that the likelihood ratio test is *equivalent* to a test based on $\bar{x}$
 
 * The p-value is $P(\bar{X} \leq 20 | \theta = 25)$ ("extreme" values of $\bar{x}$ are those that are at least as small as $\bar{x} = 20$)

 * If $H_0$ is true, then $\bar{X} \sim \text{Normal}(25, 5^2/5)$

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.4}
x_grid_1 <- seq(from = 0, to = 20, length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, 20),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 10)
)

ggplot(data = data.frame(x = c(0, 30)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = 20, size = 1.5) +
  geom_vline(data = hypothesized_vals, mapping = aes(xintercept = theta, color = Hypothesis, linetype = Hypothesis), size = 1.5) +
  coord_cartesian(xlim = c(0, 30), expand = TRUE) +
  theme_bw()
```

```{r, echo = TRUE}
pnorm(20, mean = 25, sd = 5)
```

**What about a test of $H_0: \theta = 25$ vs. $H_A: \theta = 0$?**

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.4}
x_grid_1 <- seq(from = 0, to = 20, length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, 20),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 0)
)

ggplot(data = data.frame(x = c(0, 30)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = 20, size = 1.5) +
  geom_vline(data = hypothesized_vals, mapping = aes(xintercept = theta, color = Hypothesis, linetype = Hypothesis), size = 1.5) +
  coord_cartesian(xlim = c(0, 30), expand = TRUE) +
  theme_bw()
```

```{r, echo = TRUE}
pnorm(20, mean = 25, sd = 5)
```

**What about a test of $H_0: \theta = 25$ vs. $H_A: \theta = 20$?**

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height = 1.4}
x_grid_1 <- seq(from = 0, to = 20, length = 101)

region_to_shade1 <- data.frame(
  x = c(0, x_grid_1, 20),
  y = c(0, dnorm(x_grid_1, mean = 25, sd = sqrt(5), log = FALSE), 0)
)

hypothesized_vals <- data.frame(
  Hypothesis = c("H0", "HA"),
  theta = c(25, 20)
)

ggplot(data = data.frame(x = c(0, 30)), mapping = aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 25, sd = sqrt(5))) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = 20, size = 1.5) +
  geom_vline(data = hypothesized_vals, mapping = aes(xintercept = theta, color = Hypothesis, linetype = Hypothesis), size = 1.5) +
  coord_cartesian(xlim = c(0, 30), expand = TRUE) +
  theme_bw()
```

```{r, echo = TRUE}
pnorm(20, mean = 25, sd = 5)
```

**Main point: In this framework the null hypothesis is "privileged"**

 * Everything is framed in terms of strength of evidence against the null hypothesis.
 * p-value calculations done assuming $H_0$ is true
 * In many examples (like this one), details of $H_A$ are relatively unimportant (here, only determines direction of "more extreme")
